---
title: "Characterizing and Displaying Multivariate Data"
author: "Blake Pappas"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: false
    toc_depth: 3
    fig_width: 6
    fig_height: 5
    fig_caption: yes
header-includes:
   - \usepackage{animate}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Descriptive Statistics

### Sample Covariance Visualization

```{r}
set.seed(123)
library(MASS)
dat <- mvrnorm(n = 20, mu = c(0, 0), Sigma = matrix(c(4, 1.4, 1.4, 1), 2))
n <- dim(dat)[1]
par(mar = c(3.6, 3.6, 0.8, 0.6), las = 1)
layout(matrix(c(1, 1, 2, 3), nrow = 2, ncol = 2))
plot(dat, pch = 16, las = 1, xlab = "", ylab = "")
mtext(expression(X[j]), 1, line = 2); mtext(expression(X[k]), 2, line = 2)
text(-4, 2, expression(paste(S[jk], " = ")))
text(-3.3, 2, round(cov(dat[, 1], dat[, 2]), 2))
abline(h = mean(dat[, 2]), lty = 2); abline(v = mean(dat[, 1]), lty = 2)
plot(1:n, dat[, 1], pch = 16, xlab = "", ylab = "")
abline(h = mean(dat[, 1]))
mtext(expression(X[j]), 1, line = 2)
plot(1:n, dat[, 2], pch = 16, xlab = "", ylab = "")
abline(h = mean(dat[, 2]))
mtext(expression(X[k]), 1, line = 2)
```

### Sample and Population Covariance

Here we simulate data with size sample $n = 20$ from a bivariate normal distribution with *population covariance* $\rho_{12} = 0$.
We calculate the *sample covariance* $s_{12}$ for each simulated data set, and we repeat this process 1,000 times. 

The main purpose of this exercise is to demonstrate that one can conduct a *Monte Carlo* experiment to approximate the *sampling distribution* of $s_{12}$. 

```{r}
dat <- replicate(1000, mvrnorm(n = 20, mu = c(0, 0), Sigma = matrix(c(1, 0, 0, 1), 2)))

s12 <- apply(dat, 3, function(x) cov(x[, 1], x[, 2]))
hist(s12, 20, las = 1, main = "")
abline(v = 0, col = "red")
```

### Bivariate Data Example

```{r}
data <- cbind(x1 = c(42, 52, 88, 58, 60), x2 = c(4, 5, 7, 4, 5))
(means <- apply(data, 2, mean))
cov(data)
cor(data)
```

### Generliazed Variance

```{r}
data(mtcars)
vars <- which(names(mtcars) %in% c("mpg", "disp", "hp", "drat", "wt"))
car <- mtcars[, vars]; S <- cov(car)
(genVar <- det(S))
```


## Graphs and Visualization

### pairs

```{r}
head(iris)
pairs(iris[, -5], las = 1, col = rep(c("green", "red", "blue"), each = 50), cex = 0.8)
```

### ggpairs

```{r, warning = FALSE}
library(GGally)
library(ggplot2)
p <- ggpairs(iris[, -5], aes(color = iris$Species)) + theme_bw()
# Change Color Manually
# Loop Through Each Plot, Changing Relevant Scales
for(i in 1:p$nrow) {
  for(j in 1:p$ncol) {
    p[i, j] <- p[i, j] + 
        scale_fill_manual(values = c("green", "red", "blue")) +
        scale_color_manual(values = c("green", "red", "blue"))  
  }
}
p
```

### 3D Scatterplot

```{r}
library(scatterplot3d)
scatterplot3d(iris[, 1:3], pch = 19, color = rep(c("green", "red", "blue"), each = 50), grid = TRUE, box = FALSE, mar = c(3, 3, 0.5, 3))
```


### Chernoff Faces

```{r, warning = FALSE}
library(aplpack)
par(mar = rep(0, 4))
faces(mtcars, cex = 0.8)

USArrests
faces(USArrests, cex = 0.8)
```

### Visualizing Summary Statistics

```{r}
library(ggcorrplot)

# Compute a Correlation Matrix
corr <- round(cor(car), 1)

# Visualize
ggcorrplot(corr, p.mat = cor_pmat(car),
           hc.order = TRUE, type = "lower",
           color = c("#FC4E07", "white", "#00AFBB"),
           outline.col = "white", lab = TRUE)
```