---
title: "Support Vector Machine and Neural Networks"
author: "Blake Pappas"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
bookmarks: no
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning = FALSE, message = FALSE}
library(dplyr)
library(e1071)
library(caret)
library(neuralnet)
library(FSelectorRcpp)
```

# Part I. Support Vector Machine in R

# Load the "bank_small.csv" Dataset:

```{r, warning = FALSE}
bank = read.csv("bank_small.csv", stringsAsFactors = TRUE)

train_rows = createDataPartition(y = bank$y,
                                 p = 0.80, list = FALSE)

bank_train = bank[train_rows, ]
bank_test = bank[-train_rows, ]
```


# Train a Linear SVM Model Using All Variables

```{r, warning = FALSE}
# Train the Model
svm_model_linear = svm(y ~ ., data = bank_train,
                       kernel = "linear")

# Make Predictions
pred_svm_model_linear = predict(svm_model_linear, bank_test)

# Performance Evaluation
# Confusion Matrix
confusionMatrix(pred_svm_model_linear, bank_test$y, mode = "prec_recall", positive = "yes")
```


# Train a 2-Degree Polynomial SVM Model Using All Variables

```{r, warning = FALSE}
# Train the Model
svm_model_polynomial = svm(y ~ ., data = bank_train,
                kernel = "polynomial", degree = 2)

# Make Predictions
pred_svm_model_polynomial = predict(svm_model_polynomial, bank_test)

# Performance Evaluation
# Confusion Matrix
confusionMatrix(pred_svm_model_polynomial, bank_test$y, mode = "prec_recall", positive = "yes")
```


# Train a Gaussian (Radial) SVM Model Using All Variables

```{r, warning = FALSE}
# Train the Model
svm_model_radial = svm(y ~ ., data = bank_train,
                       kernel = "radial")

# Make Predictions
pred_svm_model_radial = predict(svm_model_radial, bank_test)

# Performance Evaluation
# Confusion Matrix
confusionMatrix(pred_svm_model_radial, bank_test$y, mode = "prec_recall", positive = "yes")
```


# Train a Linear SVM Model Using the Filter Approach (Information Gain)

```{r, warning = FALSE}
# Information Gain
IG_linear = information_gain(y ~ ., data = bank_train)

# Select Top 10 Attributes
top10 = cut_attrs(IG_linear, k = 10)

bank_top10_train = bank_train %>% select(top10, y)
bank_top10_test = bank_test %>% select(top10, y)

# Train the Model
svm_model_linear_IG = svm(y ~ ., data = bank_top10_train,
                       kernel = "linear")

# Make Predictions
pred_svm_model_linear_IG = predict(svm_model_linear_IG, bank_top10_test)

# Performance Evaluation
# Confusion Matrix
confusionMatrix(pred_svm_model_linear_IG, bank_top10_test$y, mode = "prec_recall", positive = "yes")
```


# Train a 2-Degree Polynomial SVM Model Using the Filter Approach (Information Gain)

```{r, warning = FALSE}
# Information Gain
IG_polynomial = information_gain(y ~ ., data = bank_train)

# Select Top 7 Attributes
top7 = cut_attrs(IG_polynomial, k = 7)

bank_top7_train = bank_train %>% select(top7, y)
bank_top7_test = bank_test %>% select(top7, y)

# Train the Model
svm_model_polynomial_IG = svm(y ~ ., data = bank_top7_train,
                          kernel = "polynomial")

# Make Predictions
pred_svm_model_polynomial_IG = predict(svm_model_polynomial_IG, bank_top7_test)

# Performance Evaluation
# Confusion Matrix
confusionMatrix(pred_svm_model_polynomial_IG, bank_top7_test$y, mode = "prec_recall", positive = "yes")
```


# Train a Gaussian (Radial) SVM Model Using the Filter Approach (Information Gain)

```{r, warning = FALSE}
# Information Gain
IG_radial = information_gain(y ~ ., data = bank_train)

# Select Top 5 Attributes
top5 = cut_attrs(IG_radial, k = 5)

bank_top5_train = bank_train %>% select(top5, y)
bank_top5_test = bank_test %>% select(top5, y)

# Train the Model
svm_model_radial_IG = svm(y ~ ., data = bank_top5_train,
                       kernel = "radial")

# Make Predictions
pred_svm_model_radial_IG = predict(svm_model_radial_IG, bank_top5_test)

# Performance Evaluation
# Confusion Matrix
confusionMatrix(pred_svm_model_radial_IG, bank_top5_test$y, mode = "prec_recall", positive = "yes")
```


# Part II. Neural Networks in R

# Load the "wine.csv" Dataset:

```{r, warning = FALSE}
wine = read.csv("wine.csv")

train_rows = createDataPartition(y = wine$Type,
                                 p = 0.80, list = FALSE)
```

```{r, warning = FALSE}
# Data Pre-Processing
normalize = function(x) {
  return((x - min(x)) / (max(x) - min(x)))
  }

wine = wine %>%
  mutate(A = ifelse(Type == "A", 1, 0),
         B = ifelse(Type == "B", 1, 0),
         C = ifelse(Type == "C", 1, 0)) %>%
  mutate_at(2:14, normalize)

wine_train = wine[train_rows, ]
wine_test = wine[-train_rows, ]
```


# Train a Neural Network Classifier: 1 Hidden Layer with 2 Nodes and a Learning Rate of 0.1

```{r, warning = FALSE}
# Train the Neural Network
nn_model = neuralnet(A + B + C ~ Alcohol + Malic_Acid + Ash + Ash_Alcalinity + Magnesium + Total_Phenols + Flavanoids + Nonflavanoid_Phenols + Proanthocyanins + Color_Intensity + Hue + OD280_OD315 + Proline,
                     data = wine_train,
                     act.fct = "logistic",
                     linear.output = FALSE,
                     hidden = 2,
                     algorithm = "backprop",
                     learningrate = 0.1,
                     rep = 1)

# Make Predictions and Evaluate Performance
pred = neuralnet::compute(nn_model, wine_test[, 2:14])$net.result

outcomes = c("A", "B", "C")
pred_label = outcomes[max.col(pred)]
confusionMatrix(factor(pred_label), factor(wine_test$Type))
```


# Train a Neural Network Classifier: 3 Hidden Layers with 2 Nodes Each and a Learning Rate of 0.1

```{r, warning = FALSE}
# Train the Neural Network
nn_model = neuralnet(A + B + C ~ Alcohol + Malic_Acid + Ash + Ash_Alcalinity + Magnesium + Total_Phenols + Flavanoids + Nonflavanoid_Phenols + Proanthocyanins + Color_Intensity + Hue + OD280_OD315 + Proline,
                     data = wine_train,
                     act.fct = "logistic",
                     linear.output = FALSE,
                     hidden = rep(2, 3),
                     algorithm = "backprop",
                     learningrate = 0.1,
                     rep = 1)

# Make Predictions and Evaluate Performance
pred = neuralnet::compute(nn_model, wine_test[, 2:14])$net.result

outcomes = c("A", "B", "C")
pred_label = outcomes[max.col(pred)]
confusionMatrix(factor(pred_label), factor(wine_test$Type))
```


# Train a Neural Network Classifier: 5 Hidden Layers with 2, 3, 5, 4, and 2 Nodes and a Learning Rate of 0.1

```{r, warning = FALSE}
# Train the Neural Network
nn_model = neuralnet(A + B + C ~ Alcohol + Malic_Acid + Ash + Ash_Alcalinity + Magnesium + Total_Phenols + Flavanoids + Nonflavanoid_Phenols + Proanthocyanins + Color_Intensity + Hue + OD280_OD315 + Proline,
                     data = wine_train,
                     act.fct = "logistic",
                     linear.output = FALSE,
                     hidden = c(2, 3, 5, 4, 2),
                     algorithm = "backprop",
                     learningrate = 0.1,
                     rep = 1)

# Make Predictions and Evaluate Performance
pred = neuralnet::compute(nn_model, wine_test[, 2:14])$net.result

outcomes = c("A", "B", "C")
pred_label = outcomes[max.col(pred)]
confusionMatrix(factor(pred_label), factor(wine_test$Type))
```


# Train a Neural Network Classifier: 3 Hidden Layers with 2 Nodes Each and a Learning Rate of 0.01

```{r, warning = FALSE}
# Train the Neural Network
nn_model = neuralnet(A + B + C ~ Alcohol + Malic_Acid + Ash + Ash_Alcalinity + Magnesium + Total_Phenols + Flavanoids + Nonflavanoid_Phenols + Proanthocyanins + Color_Intensity + Hue + OD280_OD315 + Proline,
                     data = wine_train,
                     act.fct = "logistic",
                     linear.output = FALSE,
                     hidden = rep(2, 3),
                     algorithm = "backprop",
                     learningrate = 0.01,
                     rep = 1)

# Make Predictions and Evaluate Performance
pred = neuralnet::compute(nn_model, wine_test[, 2:14])$net.result

outcomes = c("A", "B", "C")
pred_label = outcomes[max.col(pred)]
confusionMatrix(factor(pred_label), factor(wine_test$Type))
```


# Train a Neural Network Classifier: 3 Hidden Layers with 2 Nodes Each and a Learning Rate of 0.3

```{r, warning = FALSE}
# Train the Neural Network
nn_model = neuralnet(A + B + C ~ Alcohol + Malic_Acid + Ash + Ash_Alcalinity + Magnesium
                     + Total_Phenols + Flavanoids + Nonflavanoid_Phenols + Proanthocyanins
                     + Color_Intensity + Hue + OD280_OD315 + Proline,
                     data = wine_train,
                     act.fct = "logistic",
                     linear.output = FALSE,
                     hidden = 2,
                     algorithm = "backprop",
                     learningrate = 0.1,
                     rep = 1)

# Make Predictions and Evaluate Performance
pred = neuralnet::compute(nn_model, wine_test[, 2:14])$net.result

outcomes = c("A", "B", "C")
pred_label = outcomes[max.col(pred)]
confusionMatrix(factor(pred_label), factor(wine_test$Type))
```


# Train a Neural Network Classifier: 1 Hidden Layer with 2 Nodes and a Learning Rate of 0.01

```{r, warning = FALSE}
# Train the Neural Network
nn_model = neuralnet(A + B + C ~ Alcohol + Malic_Acid + Ash + Ash_Alcalinity + Magnesium + Total_Phenols + Flavanoids + Nonflavanoid_Phenols + Proanthocyanins + Color_Intensity + Hue + OD280_OD315 + Proline,
                     data = wine_train,
                     act.fct = "logistic",
                     linear.output = FALSE,
                     hidden = 2,
                     algorithm = "backprop",
                     learningrate = 0.01,
                     rep = 1)

# Make Predictions and Evaluate Performance
pred = neuralnet::compute(nn_model, wine_test[, 2:14])$net.result

outcomes = c("A", "B", "C")
pred_label = outcomes[max.col(pred)]
confusionMatrix(factor(pred_label), factor(wine_test$Type))
```


# Train a Neural Network Classifier: 1 Hidden Layer with 2 Nodes and a Learning Rate of 0.5

```{r, warning = FALSE}
# Train the Neural Network
nn_model = neuralnet(A + B + C ~ Alcohol + Malic_Acid + Ash + Ash_Alcalinity + Magnesium + Total_Phenols + Flavanoids + Nonflavanoid_Phenols + Proanthocyanins + Color_Intensity + Hue + OD280_OD315 + Proline,
                     data = wine_train,
                     act.fct = "logistic",
                     linear.output = FALSE,
                     hidden = 2,
                     algorithm = "backprop",
                     learningrate = 0.5,
                     rep = 1)

# Make Predictions and Evaluate Performance
pred = neuralnet::compute(nn_model, wine_test[, 2:14])$net.result

outcomes = c("A", "B", "C")
pred_label = outcomes[max.col(pred)]
confusionMatrix(factor(pred_label), factor(wine_test$Type))
```