---
title: 'Multiple Linear Regression (Inference and Prediction) - Lab'
author: "Blake Pappas"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Housing Values in Suburbs of Boston

The Boston housing data was collected in 1978. Each of the 506 entries represent aggregated data about 14 features for homes from various suburbs in Boston, MA. 

*Data Source:* Harrison, D. and Rubinfeld, D.L. (1978) Hedonic prices and the demand for clean air. **J. Environ. Economics and Management** 5, 81–102.


## Load the Dataset

**Code:**

```{r}
library(MASS)
data(Boston)
```

We will use only the following variables for conducting data analysis:

1. \texttt{medv}: median value of owner-occupied homes in $\$1000s$; 

2. \texttt{lstat}: lower status of the population (percent);

3. \texttt{rm}: average number of rooms per dwelling;

4. \texttt{crim}: per capita crime rate by town

**Code:**

The code below can be used to extract these variables.

```{r}
vars <- c("medv", "lstat", "rm", "crim")
data <- Boston[, vars]
```

 
## Exploratory Data Analysis

### Numerical Summary

1. Use `summary` command to produce various numerical summaries of each of the 4 variables under consideration. 

**Code:**

```{r}
summary(data)
```

### Graphical Summary

2. Make a boxplot for each variable.

**Code:**

```{r}
par(mfrow = c(1, 4))
boxplot(data$medv, main = 'medv', ylab = 'Median Home Value ($)')
boxplot(data$lstat, main = 'lstat', ylab = 'Lower Status of Population (%)')
boxplot(data$rm, main = 'rm', ylab = 'Average Rooms per Dwelling')
boxplot(data$crim, main = 'crim', ylab = 'Per Capita Crime Rate by Town')
```

3. Briefly discuss the shape of the distribution of each variable.

**Code:**

```{r}
par(mfrow = c(1, 4))
hist(data$medv, main = 'medv', xlab = '', ylab = 'Median Home Value ($)')
hist(data$lstat, main = 'lstat', xlab = '', ylab = 'Lower Status of Population (%)')
hist(data$rm, main = 'rm', xlab = '', ylab = 'Average Rooms per Dwelling')
hist(data$crim, main = 'crim', xlab = '', ylab = 'Per Capita Crime Rate by Town')
```

**Answer: The variable `mdev` is positively skewed, asymmetric, and unimodal. The variable `lstat` is positively skewed, asymmetric, and unimodal. Although it may appear normally distributed, the variable `rm` is slightly positively skewed, asymmmetric, and unimodal. The variable `crim` is positively skewed, asymmetric and unimodal.**


4. Create a scatterplot matrix to explore the interdependence between these variables.

**Code:**

```{r, warning = FALSE}
pairs(data, cex = 0.95, col = "blue", las = 1)
library(ggplot2)
library(GGally)
ggpairs(data)
```

## Model Fitting 

Here we will use `medv` as the response and `lstat`, `rm`, `crim` as the predictors.  

### Simple Linear Regression

5. Fit a simple linear regression.

**Code:**

```{r}
# Simple Linear Regression Using mdev as the Response and rm as the Predictor
slr <- lm(medv ~ rm, data = data) # This will be the reduce model for the F-test
summary(slr)
```

```{r}
y <- data$medv; x <- data$rm
y_diff <- y - mean(y)
x_diff <- x - mean(x)
beta_1 <- sum(y_diff * x_diff) / sum((x_diff)^2)
beta_1

beta_0 <- mean(y) - mean(x) * beta_1
beta_0

par(las = 1, mar = c(4.1, 4.1, 1.1, 1.1))
plot(x, y, pch = 16, xlab = "Average Rooms per Dwelling (rm)", ylab = "Median Home Value ($) (medv)")
grid()
abline(a = beta_0, b = beta_1, col = "red")
```

6. Write down the fitted linear regression equation.

**Answer: `medv` = -34.671 + 9.102 x `rm` + $\epsilon$, where Y = `medv`, X = `rm`, $\hat{\beta_0}$ = -34.671, $\hat{\beta_1}$ = 9.102, and $\epsilon$ = stochastic error.**



### Multiple Linear Regression

7. Fit a multiple linear regression using all predictors.

**Code:**

```{r}
mlr <- lm(medv ~ ., data = data) # This will be the full model for the F-test
summary(mlr)
```
8. Write down the fitted linear regression equation.

**Answer: `medv` = -2.56225 - 0.57849 x `lstat` + 5.21695 x `rm` - 0.10294 x `crim` + $\epsilon$**



9. Perform an overall F-test and state the hypotheses, test statistic, p-value, decision, and conclusion.

**Code:**

```{r}
anova(slr, mlr)
```

**Answer: The null hypothesis was H0 : ${\beta}$lstat = ${\beta}$rm = ${\beta}$crim = 0. The alternative hypothesis was HA : at least one of the three coefficients $\neq$ 0. The test statistic was 115.05, the p-value was 2.2e-16, and conclusion was to reject the null hypothesis.**