---
title: "Multiple Linear Regression (Model Selection and Model Checking) - Lab"
author: "Blake Pappas"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Savings Rates in 50 Countries

The savings data frame has 50 rows (countries) and 5 columns (variables):

1. \texttt{sr}: savings rate - personal saving divided by disposable income
*This variable will be used as the response*
2. \texttt{pop15}: percent population under age of 15
3. \texttt{pop75}: percent population over age of 75
4. \texttt{dpi}: per-capita disposable income in dollars
5. \texttt{ddpi}: percent growth rate of dpi

The data is averaged over the period 1960-1970. 



*Data Source:* Belsley, D., Kuh. E. and Welsch, R. (1980) *Regression Diagnostics* Wiley.


Load the dataset.

**Code:**

```{r}
data(savings, package = "faraway")
head(savings)
```

1. Perform the best subset selection and select the "best" model using $R^{2}_{adj}$.

**Code:**

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(caret)
library(leaps)
models <- regsubsets(sr ~ ., data = savings) # regsubsets = the function for model selection
summary(models) # Gives best model based on the number of predictors


res.sum <- summary(models)

criteria <- data.frame(
  Adj.R2 = res.sum$adjr2,
  Cp = res.sum$cp,
  BIC = res.sum$bic)

criteria

# Plot of Adjusted R-Squared
plot(2:5, criteria$Adj.R2, las = 1, xlab = "p", ylab = "", pch = 16, col = "gray",
     main = expression(R['adj']^2))
points(4, criteria$Adj.R2[3], col = "blue", pch = 16)
```

**Answer: The best model using $R^{2}_{adj}$ is the third model, which uses the `pop15`, `pop75`, and `ddpi` as the predictors.**


2. Perform a stepwise selection using *AIC*.

**Code:**

```{r}
full <- lm(sr ~ ., data = savings)
step(full, direction = "both")
```

**Answer: Using the stepwise selection, we are brought to an AIC of 136.45, which uses `pop15`, `pop75`, and `ddpi` as the predictors for the regression having the best goodness of fit.**

3. Perform a general linear F-test (with $\alpha = 0.1$) to choose between the full model (i.e., using the all 4 predictors) and the reduced model that includes `pop15`, `pop75`, and `ddpi` as the predictors.

**Code:**

```{r}
# Reduced Model
reduce <- lm(sr ~ pop15 + pop75 + ddpi, data = savings)
summary(reduce)

# Full Model
full <- lm(sr ~ ., data = savings)
summary(full)

# General Linear F-Test
anova(reduce, full)
```

**Answer: Since the p-value of this general linear F-test is greater than α, we fail to reject H0 and conclude that we do not have sufficient
evidence to support that at least one of the three regression coefficients is not equal to 0.**

4. Make a residual plot of the model selected by *AIC* and comment on the model assumptions.

**Code:**

```{r}
mod <- lm(formula = sr ~ pop15 + pop75 + ddpi, data = savings)
plot(mod$fitted.values, mod$residuals, pch = 16, col = "blue", xlab = "Fitted Values", ylab = "Residuals", main = "Residuals vs. Fitted")
abline(h = 0, col = "red")
```

**Answer: There is no major concern with the model assumptions, as the plot of the residuals appears to be random.**

5. Use both a histogram and qqplot to examine the normality assumption on error.

**Code:**

```{r}
# Histogram
(sd <- sd(mod$residuals))

par(las = 1)
hist(mod$residuals, 5, prob = T, col = "lightblue", border = "gray", main = "Histogram of Residuals", xlab = "Residuals")
xg <- seq(-10, 10, 1)
yg <- dnorm(xg, 0, sd)
lines(xg, yg)

# qqplot
qqnorm(mod$residuals, pch = 16, las = 1, col = "gray", xlab = "Normal Quantiles", ylab = "Residuals")
qqline(mod$residuals)
```

**Answer: There is no major concern regarding normality with this model. The distribution of the residuals appear to be approximately normally distributed. The Normal Q-Q Plot appears to run closely (in an S-shaped pattern) to the trend line,with little deviation.**

6. Calculate the leverage values to check if there is any high leverage points (i.e., $h > \frac{2p}{n}$).

**Code:**

```{r}
step_savings <- step(full, trace = F) # Trace = full
X <- model.matrix(step_savings) # Model Design Matrix
H <- X %*% solve((t(X) %*% X)) %*% t(X)
lev <- hat(X) # Calculates leverage
high_lev <- which(lev >= 2 * 3 / 30) # Finds the high leverage values
attach(savings)

# Plot of Leverage Points
par(las = 1)
plot(pop15, dpi, cex = sqrt(5 * lev), col = "blue", ylim = c(0, 5000))
points(pop15[high_lev], dpi[high_lev], col = "red", pch = 16,
       cex = sqrt(5 *lev[high_lev]))
```

**Answer: There are two high leverage points between `pop15` and `dpi`. They exist in observations 23 an 49.**

7. Compute jackknife residuals to identify outlier(s).

**Code:**

```{r}
jack <- rstudent(step_savings)

par(las = 1)
plot(jack, pch = 16, cex = 0.8, col = "blue", main = "Jackknife Residuals",
     xlab = "", ylab = "")
abline(h = 0, lty = 2, col = "gray")
```

**Answer: Looking at the graph, there do not appear to be many outliers.**

8. Identifying influential observations by computing DFFITS.

**Code:**

```{r, message = FALSE, warning = FALSE}
library(olsrr)
ols_plot_dffits(step_savings)
```

**Answer: There are three influential observations for `sr`: Observations 23, 46, and 49.**