---
title: "Class Probability Prediction"
author: "Blake Pappas"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
bookmarks: no
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Class Probability Prediction in R
# Load the following packages:

```{r, warning = FALSE, message = FALSE}
library(dplyr)
library(caret)
library(e1071)
library(pROC)
```

# In this exercise, we use the "bank.csv" file. The goal is to predict outcome "y" - whether a customer makes a deposit as a result of the bank's marketing activity. Note: the delimiter of this .csv file is not comma, but ";"

# P1: Import the dataset:

```{r}
bank = read.csv("bank.csv", sep = ";")
```

# P2: Split the dataset into 75% training and 25% testing
```{r}
library(caret)

train_rows = createDataPartition(y = bank$y,
                                 p = 0.75, list = FALSE)

bank_train = bank[train_rows, ]

bank_test = bank[-train_rows, ]
```

# P3: Build a naive Bayes model
```{r}
library(e1071)

NB_model = naiveBayes(y ~ ., data = bank_train)
```

# P4: Evaluate the performance of your model. Report AUC.
```{r}
# Make Categorical Predictions
pred_nb = predict(NB_model, bank_test, type = "class")

# Make Class Probability Predictions
prob_pred_nb = predict(NB_model, bank_test, type = "raw")

# Evaluate Model Performance
confusionMatrix(pred_nb, as.factor(bank_test$y),
                mode = "prec_recall", positive = "yes")

# Make ROC Curve for Class “yes”
library(pROC)

roc_nb = roc(response = ifelse(bank_test$y == "yes", 1, 0),
             predictor = prob_pred_nb[, 2])

# Calculate AUC
auc(roc_nb)
```

# P5: Plot the ROC curve
```{r}
plot(roc_nb)
```

# P6: Plot the lift curve. What is the lift ratio for the top 20% of customers? What does that mean?
```{r}
bank_test_lift = bank_test %>%
  mutate(prob = prob_pred_nb[, 2]) %>%
  arrange(desc(prob)) %>%
  mutate(y_yes = ifelse(y == "yes", 1, 0)) %>%
  # The Following Two Lines Make the Lift Curve
  mutate(x = row_number() / nrow(bank_test),
         y = (cumsum(y_yes) / sum(y_yes)) / x)

plot(bank_test_lift$x, bank_test_lift$y)
```

# Answer: The lift ratio for the top 20% of customers is about three. This means that the model's top 20% of customers are about three times as good at predicting the outcome than random guesses.