---
title: 'Introduction to R'
author: "Blake Pappas"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: false
    toc_depth: 3
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Basic

### What is in the Workspace?
```{r}
ls()
```

### Where is My \texttt{R} Working Directory?

```{r}
getwd()
```


### What is in My Working Directory?

```{r}
dir()
```

### "$c$" Combines Sets of Numbers (or Datasets)

```{r}
x <- c(2, 3, 20)
```

*Note*: \texttt{R} is case sensitive. Type "X" in \texttt{R} console and then click "Enter" to see what happens.

### Now Re-Check Workspace

```{r}
ls()

# Print Out x
x
```

### Reassign "x" to Another Name

```{r}
x2 <- x
ls()
```

### Remove "x" and Create Another Object "x3"

```{r}
rm(x)
x3 <- c(3, 4, 5)
```


### Question: How Would You Combine "x2" and "x3" to Make a New Data Set ?

```{r}
x4 <- x2 + x3
x4
```

## Arithemtic

### Add Numbers in \texttt{R}

```{r}
A <- 2
B <- 10
Y <- A + B

A <- c(2, 3, 4)
B <- c(10, 100, 1000)
Y <- A + B
Y # Note that the numbers have been added row by row like a spreadsheet
```


### Some Other Operations

```{r}
2^4
2 * (1 + 4)
sqrt(81)
exp(2)
```

### Generating a Sequence 

```{r}
1:10
-5:5
```

*Question*: How Would You Generate the Values in Order from 5 to 1?
```{r}
5:1
```

### Generating Sin Wave

```{r}
x <- 0:100

# Hundred Values Between 0 and 2 * pi
x <-  2 * pi * (x / 100)
y <- sin(x)

# Plot the Sin Wave 
plot(x, y)

# Change the Plot to Connect Points with a Line Instead of Points
plot(x, y, type = "l")

# Another Way of Creating the x
x <- seq(0, 2 * pi, length.out = 101)
```


### Use \texttt{R} to Generate Random Values 

```{r}
# Generating 10 Random Numbers Between 0 and 1 
U <- runif(10)
U

# Generating 30 Random Numbers from a Standard Normal Distribution
V <- rnorm(n = 30, mean = 0, sd = 1)
hist(V)
qqnorm(V)
qqline(V)
```

## Subsetting

### Load a Data Set 

```{r}
load("BT.RData") # Uses CWD

# Copy for Easy Typing
BT <- BoulderJuneTemperature$Temp
BAll <- BoulderJuneTemperature
head(BT)
head(BAll, 10)
```


### Print the First 10 Values
```{r}
BT[1:10]
```

### An Indicator for All Values Over 70
```{r}
ind70 <- BT > 70
ind70
```

 
```{r}  
# Temperatures with Values Over 70
BT[ind70]
```

```{r}  
# The Years with Values Over 70
BAll$Year[ind70]
```

*Question*: How many years exceed 70 degrees?

### Working with This Data as a Matrix

```{r}
dim(BAll)

# This Is the First Row and First Column
BAll[1, 1]

# First Row
BAll[1,]

# First Column
BAll[, 1]

# Column with Year (This Is Also Column 1)
BAll[, "Year"]

# Second Column (You Could Also Use "Temp" to Refer to This)
BAll[, 2]

# Rows 10 Through 20
BAll[10:20,]
```

*Exercise*: Plot the temperatures by year
  
## Apply Functions in \texttt{R}

### *apply* Functions

1. a family of functions in R which allow you to repetitively perform an action on multiple chunks of data

2. run faster than loops and often require less code. 

Let's take a look at some examples

### Load the Boulder Temperature Data Set Into R 
```{r}
load("BoulderTemperature.RData") # Monthly mean temperatures #Uses CWD
dim(BoulderTemperature)

# Check Out First Row
BoulderTemperature[1,]

# Extract 1991 - 2010
yr <- rownames(BoulderTemperature)
index <- which(yr %in% 1991:2010) 
tempData <- BoulderTemperature[index,]

# Check This Out 
tempData
```

### The "*apply*" Function

```{r}
# Means by Rows of This Table
byYear <- apply(tempData, 1, FUN = mean) # By rows, 1 = first index
byYear
#rowMeans(tempData) Will Also Do the Same

# Means by Columns
byMonth <- apply(tempData, 2, FUN = mean) # By cols, 2 = second index
byMonth
#colMeans(tempData) Will Also Do the Same
```


## Writing Functions in \texttt{R}


### Finding the Interquartile Range (IQR)

```{r}
# 75% Quantile  
BT75 <- quantile(BT, 0.75)

# Question: Find the Interquartile Range 75% - 25% Quantiles and Check This Against the Built-In Function
IQR(BT)
```

### Building Your Own Function

Here is a function that adds the squares of two numbers. It has three parts, the *calling arguments*, the *body* where you do the work and then *returning any results*.

```{r}
myFun <- function(a, b) {
    result <- a^2 + b^2
    return(result)
}

test1 <- myFun(2, 3)
test1
test2 <- myFun(1:5, 11:15)
test2
```

Note that the "a", "b" and result are only used inside the function and do not appear in your workspace.
Also since the body is normal R code, this works for vectors automatically.

### Building Your Own IQR Function

```{r}
myIQR <- function(y) {
  IQR <- quantile(y, 0.75, names = FALSE) - quantile(y, 0.25, names = FALSE)
  return(IQR)
}
myIQR(BT)
```

### Modify This Function to Work with NAs
```{r}
test <- c(BT, NA)  
myIQR <- function(y, na.rm = FALSE) {
            IQR <- quantile(y, 0.75, names = FALSE, na.rm = na.rm) - 
                   quantile(y, 0.25, names = FALSE, na.rm = na.rm) 
            return(IQR)
}
myIQR(test, na.rm = T)
```

### Adding Warning Message 

```{r}
myIQR <- function(y, na.rm = FALSE) {
# An Example of Adding a Warning     
    if(na.rm) {
        warning("NAs removed from the data")
    }
            IQR <- quantile(y, 0.75, names = FALSE, na.rm = na.rm) - 
                   quantile(y, 0.25, names = FALSE, na.rm = na.rm) 
            return(IQR)
} 
myIQR(test, na.rm = T)
```