---
title: "Support Vector Machine"
author: "Blake Pappas"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
bookmarks: no
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Support Vector Machine in R
# Load the following packages:
```{r, warning = FALSE, message = FALSE}
library(caret)
library(dplyr)
library(e1071)
```

# In this exercise, we use the "credit.csv" file. This file concerns credit card applications.
# All attribute names and values have been changed to meaningless symbols to protect confidentiality of the data.
# The outcome variable is A16, attributes are A1 - A15.


# P1: Import the dataset. Split it to 80% training and 20% testing.
```{r}
credit = read.csv("credit.csv", stringsAsFactors = TRUE)

train_rows = createDataPartition(y = credit$A16,
                                 p = 0.80, list = FALSE)

credit_train = credit[train_rows, ]
credit_test = credit[-train_rows, ]
```


# Build a SVM model with a linear kernel
```{r}
svm_model_linear = svm(A16 ~ ., data = credit_train,
                kernel = "linear")

# Make Predictions
pred_svm_model_linear = predict(svm_model_linear, credit_test)

# Performance Evaluation
# Confusion Matrix
confusionMatrix(pred_svm_model_linear, credit_test$A16, mode = "prec_recall", positive = "+")
```


# Build a SVM model with a radial kernel
```{r}
svm_model_radial = svm(A16 ~ ., data = credit_train,
                kernel = "radial")

# Make Predictions
pred_svm_model_radial = predict(svm_model_radial, credit_test)

# Performance Evaluation
# Confusion Matrix
confusionMatrix(pred_svm_model_radial, credit_test$A16, mode = "prec_recall", positive = "+")
```

# Use 5-fold cross validation to evaluate each model. Find the model with the best f-measure for class "+". Report the model and the f-measure.
```{r}
# Linear Kernel SVM Model
cv = createFolds(y = credit$A16, k = 5)

rmse_cv = c()

for (test_row in cv) {
  
  credit_train = credit[-test_row, ]
  credit_test = credit[test_row, ]
  
  svm_model_linear = svm(A16 ~ ., data = credit_train,
                         kernel = "linear")
  
  pred_svm_model_linear = predict(svm_model_linear, credit_test)
  
  cm = confusionMatrix(pred_svm_model_linear, credit_test$A16, mode = "prec_recall", positive = "+")
  
  print(cm$byClass[7])
}
```

```{r}
# Radial Kernel SVM Model
cv = createFolds(y = credit$A16, k = 5)

rmse_cv = c()

for (test_row in cv) {
  
  credit_train = credit[-test_row, ]
  credit_test = credit[test_row, ]
  
  svm_model_radial = svm(A16 ~ ., data = credit_train,
                         kernel = "radial")
  
  pred_svm_model_radial = predict(svm_model_radial, credit_test)
  
  cm = confusionMatrix(pred_svm_model_radial, credit_test$A16, mode = "prec_recall", positive = "+")
  
  print(cm$byClass[7])
}
```