---
title: 'COVID-19: Data Exploration'
author: "Blake Pappas"
date: 'Sep. 22, 2022'
output:
  pdf_document:
    toc: false
    toc_depth: 3
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## COVID-19 Data

Load the data using the *COVID19* R package:

```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(ggplot2)
# install.packages("COVID19")
library(COVID19)

# Load State-Level Data
raw <- covid19(c("US"), level = 2, verbose = FALSE)
```


**Question:** Pick 4 variables from the data set *raw* and explain why you picked these 4.

**Code:**
```{r}
COVID19_DATA <- select(raw, confirmed, deaths, date, people_fully_vaccinated)
```

**Answer: I have selected the following four variables from the COVID19 data set: *confirmed*, *deaths*, *date*, and *people_fully_vaccinated*. I picked these four variables because I would like to conduct a time-series analyses of rolling full vaccinations by day, deaths by day, full vaccinations by day, and death rate by day. Visualizing full vaccinations by day requires the *date* and *people_fully_vaccinated* variables. Visualizing death rates by day requires the *date*, *confirmed* and *deaths* variables. Visualizing deaths by day requires the *date* and *deaths* variables. Visualizing confirmed cases by day requires the *date* and *confirmed* variables.**


**Question:** Calculate summary statistics for these 4 variables.

**Code:**

```{r}
summary(COVID19_DATA$confirmed)
summary(COVID19_DATA$deaths)
summary(COVID19_DATA$date)
summary(COVID19_DATA$people_fully_vaccinated)
```

**Question:** Visualize these 4 variables.

**Code:**

```{r warning = FALSE}
# Rolling Sum of Full Vaccinations by Day
ggplot(data = COVID19_DATA, aes(x = date, y = people_fully_vaccinated)) +
               geom_line() +
  labs(title = 'Rolling Sum of Full COVID-19 Vaccinations by Day', x = 'Date', y = 'Full Vaccinations')
```

```{r warning = FALSE}
# COVID-19 Death Rate by Day
COVID19_DATA <- COVID19_DATA %>%
  mutate(death_rate = (deaths / confirmed) * 100)

ggplot(data = COVID19_DATA, aes(x = date, y = death_rate)) +
               geom_line() +
    labs(title = 'COVID-19 Death Rate by Day', x = 'Date', y = 'Death Rate (%)')
```

```{r warning = FALSE}
# Rolling Sum of Confirmed Cases by Day
ggplot(data = COVID19_DATA, aes(x = date, y = confirmed)) +
               geom_line() +
  labs(title = 'Rolling Sum of Confirmed COVID-19 Cases by Day', x = 'Date', y = 'Confirmed Cases')
```

```{r warning = FALSE}
# Rolling Sum of Deaths by Day
ggplot(data = COVID19_DATA, aes(x = date, y = deaths)) +
               geom_line() +
  labs(title = 'Rolling Sum of COVID-19 Deaths by Day', x = 'Date', y = 'Deaths')
```

**Question:** Briefly describe what you have learned from this data set.

**Answer: Since the first half of 2020, the death rate from COVID-19 has decreased significantly. The rolling sum of full vaccinations increased exponentially in the year 2021 but has leveled off in 2022. There has been a gradual increase in COVID-19 deaths throughout the course of the pandemic. Throughout the pandemic, confirmed COVID-19 cases have increased gradually during the spring, summer, and fall months but significantly during the winter months.**