---
title: 'Matrix Algebra - Lab'
author: "Blake Pappas"
date: 'Sep 15, 2022'
output:
  pdf_document:
    toc: false
    toc_depth: 3
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


## Motor Trend Car Road Tests

The data was extracted from the 1974 Motor *Trend US* magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles.

### Load the Dataset

**Code:**

```{r}
data(mtcars)
```


**Question:** What is the sample size of this data set and how many variables are there?

```{r}
# Sample Size
(sample_size <- dim(mtcars)[1]) 

# Number of Variables
(num_variables <- dim(mtcars)[2]) 
```

**Answer:** The sample size of this data set is 32 and there are 11 variables.

### Subset the Data

We will focus on the following variables:

1. mpg:	Miles/gallon
2. disp: Displacement (cu.in.)
3. hp:	Gross horsepower
4. drat:	Rear axle ratio
5. wt:	Weight (1000 lbs)

**Question:** Create a new data set from *mtcars* that only contains these 5 variables.   

**Code:**

```{r}
vars <- which(names(mtcars) %in% c("mpg", "disp", "hp", "drat", "wt"))
(cars <- mtcars[, vars])
```

**Question:** Compute the sample mean vector and the sample covariance matrix for this new data set. 

**Code:**

```{r}
# Sample Mean Vector
n <- dim(cars)[1]
p <- dim(cars)[2]
X <- as.matrix(cars)
ones <- rep(1, n) # Creates a vector of 1's with a length of n
(meanCal <- (1 / n) * t(X) %*% ones)

# Sample Covariance Matrix
(S <- cov(cars))
```


### Plot the Data

Please summarize this new data set using only one graph that contains most of the information.

**Code:**

```{r}
# Chernoff Faces
library(aplpack)
par(mar = rep(0, 4))
faces(cars, cex = 0.8)
```


## Multivariate Normal Distributions

In this section, we will simulate data from bivariate normal distributions with different correlation coefficients. Let's start with $\rho = 0$, that is, these two variables are independent to each other.

**Code:**

```{r, message = FALSE}
library(MASS)
n = 1000
rho <- 0
sigma <- 1
Sigma = matrix(c(sigma, sigma * rho,
                 sigma * rho, sigma), 2)
x1 <- mvrnorm(n = n, mu = c(0, 0), Sigma = Sigma)
```

### Plot This Simulated Data Set

**Code:**

```{r}
par(mar = c(3.6, 3.6, 0.8, 0.6), las = 1)
layout(matrix(c(1, 1, 2, 3), nrow = 2, ncol = 2))
plot(x1, pch = 16, las = 1, xlab = "", ylab = "")
mtext(expression(X[j]), 1, line = 2); mtext(expression(X[k]), 2, line = 2)
text(-4, 2, expression(paste(S[jk], " = ")))
text(-3.3, 2, round(cov(x1[, 1], x1[, 2]), 2))
abline(h = mean(x1[, 2]), lty = 2); abline(v = mean(x1[, 1]), lty = 2)

plot(1:n, x1[, 1], pch = 16, xlab = "", ylab = "")
abline(h = mean(x1[, 1]))
mtext(expression(X[j]), 1, line = 2)

plot(1:n, x1[, 2], pch = 16, xlab = "", ylab = "")
abline(h = mean(x1[, 2]))
mtext(expression(X[k]), 1, line = 2)
```

### Compute the Eigenvalues of the Sample Covariance Matrix

**Code:**

```{r}
eigen <- eigen(S)

# Eigenvalues
(S %*% eigen$vectors[, 1] / eigen$vectors[, 1])

# Eigenvectors
t(eigen$vectors[, 1]) %*% eigen$vectors[, 1]
t(eigen$vectors[, 2]) %*% eigen$vectors[, 2]
t(eigen$vectors[, 3]) %*% eigen$vectors[, 3]
t(eigen$vectors[, 4]) %*% eigen$vectors[, 4]
t(eigen$vectors[, 5]) %*% eigen$vectors[, 5]
```

**Question:** What are the theoretical eigenvalues and eigenvectors?

**Answer:** The theoretical eigennvectors are the direction in which the eigenvalues are stretched. The theoretical eigenvalues are the factor by which the eigenvectors are stretched.

### Repeat the Exercises Above but with $\rho = 0.9$
```{r, message = FALSE}
library(MASS)
n = 1000
rho <- 0.9
sigma <- 1
Sigma = matrix(c(sigma, sigma * rho,
                 sigma * rho, sigma), 2)
x1 <- mvrnorm(n = n, mu = c(0, 0), Sigma = Sigma)
```

```{r}
par(mar = c(3.6, 3.6, 0.8, 0.6), las = 1)
layout(matrix(c(1, 1, 2, 3), nrow = 2, ncol = 2))
plot(x1, pch = 16, las = 1, xlab = "", ylab = "")
mtext(expression(X[j]), 1, line = 2); mtext(expression(X[k]), 2, line = 2)
text(-4, 2, expression(paste(S[jk], " = ")))
text(-3.3, 2, round(cov(x1[, 1], x1[, 2]), 2))
abline(h = mean(x1[, 2]), lty = 2); abline(v = mean(x1[, 1]), lty = 2)

plot(1:n, x1[, 1], pch = 16, xlab = "", ylab = "")
abline(h = mean(x1[, 1]))
mtext(expression(X[j]), 1, line = 2)

plot(1:n, x1[, 2], pch = 16, xlab = "", ylab = "")
abline(h = mean(x1[, 2]))
mtext(expression(X[k]), 1, line = 2)
```