---
title: "Text Mining"
author: "Blake Pappas"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
bookmarks: no
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Text Mining in R
# Load the following packages:
```{r, warning = FALSE, message = FALSE}
library(tm)
library(wordcloud)
```

# In this exercise, we use the "superbowl.csv" file. It contains 2000 tweets about the Super Bowl.

# Import the dataset as a corpus:
```{r, warning = FALSE}
superbowl_text = read.csv("superbowl.csv")

superbowl_corpus = Corpus(VectorSource(superbowl_text$Tweet))
```


# Examine the 100th tweet in this corpus. What does it say?
```{r, warning = FALSE}
superbowl_corpus[[100]]$content
```


# Pre-Processing: Remove Punctuation
```{r, warning = FALSE}
superbowl_corpus = tm_map(superbowl_corpus, removePunctuation)
```


# Pre-Processing: Lower-Casing
```{r, warning = FALSE}
superbowl_corpus = tm_map(superbowl_corpus, tolower)
```


# Pre-Processing: Remove Stopwords
```{r, warning = FALSE}
superbowl_corpus = tm_map(superbowl_corpus, removeWords,
                     stopwords('english'))
```


# Pre-Processing: Word Stemming
```{r, warning = FALSE}
superbowl_corpus = tm_map(superbowl_corpus, stemDocument)
```


# Pre-Processing: Remove Excessive Blank Spaces
```{r, warning = FALSE}
superbowl_corpus = tm_map(superbowl_corpus, stripWhitespace)
```


# Obtain the Term-Document Matrix:
```{r}
dtm = DocumentTermMatrix(superbowl_corpus)

dtm_matrix = as.matrix(dtm)
```


# Find the top 5 most popular words in these tweets:
```{r}
word_freq = colSums(as.matrix(dtm))

word_freq_sorted = sort(word_freq, decreasing = TRUE)

word_freq_sorted[1:5]
```


# Now look at top 10 most popular words.
# Do you see anything unusual?
```{r}
word_freq_sorted[1:10]
```

# Answer: Yes, I see something unusual. The tenth most popular word isn't a word. It's more of an error message called "00000unknownunknownunknownunknownunknown".


# Run the following lines of code to get rid of the unusual word:
```{r, warning = FALSE}
superbowl_corpus = tm_map(superbowl_corpus, removeWords, 
                          "00000unknownunknownunknownunknownunknown")
```


# Update the word_freq variable:
```{r}
dtm = DocumentTermMatrix(superbowl_corpus)

dtm_matrix = as.matrix(dtm)

word_freq = colSums(as.matrix(dtm))

word_freq_sorted = sort(word_freq, decreasing = TRUE)

word_freq_sorted[1:10]
```


# Plot a WordCloud of top 50 most popular words.
# Fill in the first two parameters: "words" and "freq".
# The last two parameters specify the size of words to display and how many words to show.
```{r}
wordcloud(words = names(word_freq_sorted),
          freq = word_freq_sorted,
          scale = c(2, 0.25), 
          max.words = 50)
```