---
title: "Time Series Analysis 1"
author: "Blake Pappas"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: no
    toc_depth: '3'
  html_document:
    toc: no
    toc_depth: 3
    fig_width: 6
    fig_height: 4
    fig_caption: yes
header-includes: \usepackage{animate}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Time Series Data

### Lake Huron Time Series

Annual measurements of the level of Lake Huron in feet.

```{r}
par(mar = c(3.2, 3.2, 0.5, 0.5), mgp = c(2, 0.5, 0), bty = "L")
data(LakeHuron)
plot(LakeHuron, ylab = "Depth (in Feet)", xlab = "Year", las = 1)
grid()
```

### $\mathrm{CO}_{2}$ Concentration

Atmospheric concentrations of $\mathrm{CO}2$ are expressed in parts per million (ppm) and reported in the preliminary 1997 SIO manometric mole fraction scale.

```{r}
data(co2)
par(mar = c(3.8, 4, 0.8, 0.6))
plot(co2, las = 1, xlab = "", ylab = "")
mtext("Time (in Years)", side = 1, line = 2)
mtext(expression(paste("CO"[2], " Concentration (ppm)")), side = 2, line = 2.5)

# Seasonal and Trend Decomposition Using Loess (STL)
par(mar = c(4, 3.6, 0.8, 0.6))
stl <- stl(co2, s.window = "periodic")
plot(stl, las = 1)
```

### U.S. Monthly Unemployment Rates

```{r, message = FALSE, warning = FALSE}
# install.packages("quantmod")
library(quantmod)
getSymbols("UNRATE", src = "FRED")
head(UNRATE); tail(UNRATE)
chartSeries(UNRATE) 
```


## ARMA: ACF and PACF

```{r}
set.seed(123)
n = 200
WN <- rnorm(n)
par(mfrow = c(4, 2), mar = c(3.6, 3.6, 1.2, 0.6))
plot(1:n, WN, type = "l", las = 1, xlab = "", ylab = "")
mtext("WN")
acf(WN, xlab = "", ylab = "", main = "", las = 1)
mtext("ACF")

# AR(1) phi = 0.8
set.seed(123)
AR <- arima.sim(n = n, model = list(ar = 0.8))
plot(1:n, AR, type = "l", las = 1, xlab = "", ylab = "")
mtext("AR(1)")
acf(AR, xlab = "", ylab = "", main = "", las = 1)

# MA(1) theta = 0.5
set.seed(123)
MA <- arima.sim(n = n, model = list(ma = 0.5))
plot(1:n, MA, type = "l", las = 1, xlab = "", ylab = "")
mtext("MA(1)")
acf(MA, xlab = "", ylab = "", main = "", las = 1)

# ARMA(1, 1) phi = 0.8, theta = 0.5
set.seed(123)
ARMA <- arima.sim(n = n, model = list(ar = 0.8, ma = 0.5))
plot(1:n, ARMA, type = "l", las = 1, xlab = "", ylab = "")
mtext("ARMA(1,1)")
mtext("Time", side = 1, line = 2)
acf(ARMA, xlab = "", ylab = "", main = "", las = 1)
mtext("Lag", side = 1, line = 2)

par(mfrow = c(4, 2), mar = c(3.6, 3.6, 1.2, 0.6))
plot(1:n, WN, type = "l", las = 1, xlab = "", ylab = "")
mtext("WN")
pacf(WN, xlab = "", ylab = "", main = "", las = 1)
mtext("PACF")

# AR(1) phi = 0.8
set.seed(123)
AR <- arima.sim(n = n, model = list(ar = 0.8))
plot(1:n, AR, type = "l", las = 1, xlab = "", ylab = "")
mtext("AR(1)")
pacf(AR, xlab = "", ylab = "", main = "", las = 1)

# MA(1) theta = 0.5
set.seed(123)
MA <- arima.sim(n = n, model = list(ma = 0.5))
plot(1:n, MA, type = "l", las = 1, xlab = "", ylab = "")
mtext("MA(1)")
pacf(MA, xlab = "", ylab = "", main = "", las = 1)

# ARMA(1, 1) phi = 0.8, theta = 0.5
set.seed(123)
ARMA <- arima.sim(n = n, model = list(ar = 0.8, ma = 0.5))
plot(1:n, ARMA, type = "l", las = 1, xlab = "", ylab = "")
mtext("ARMA(1,1)")
mtext("Time", side = 1, line = 2)
pacf(ARMA, xlab = "", ylab = "", main = "", las = 1)
mtext("Lag", side = 1, line = 2)
```

## Lake Huron Case Study

### De-Trend 

```{r}
## Let Us Create a 'years' Variable
years <- time(LakeHuron)

## Plot Time Series
plot(LakeHuron, ylab = "Depth (in Feet)", xlab = "Year", las = 1)
grid()

## ACF and PACF
par(mfrow = c(2, 1), mar = c(4, 4, 1, 1))
acf(LakeHuron, xlab = "Lag (in Years)", ylab = "Sample ACF", main = "")
pacf(LakeHuron, xlab = "Lag (in Years)", ylab = "Sample PACF", main = "")

# Estimate the Linear Trend
lm <- lm(LakeHuron ~ years)
par(mfrow = c(2, 1), mar = c(3.5, 3.5, 1, 0.6))
plot(LakeHuron, ylab = "", xlab = "", las = 1); grid()
abline(lm, col = "blue", lty = 2)
mtext("Depth (in Feet)", 2, line = 2.4)
deTrend <- resid(lm)
plot(1875:1972, deTrend, type = "l", ylab = "", xlab = "", las = 1); grid()
abline(h = 0, col = "blue", lty = 2)
mtext("Year", 1, line = 2)
mtext("Depth (in Feet)", 2, line = 2.4)

## ACF and PACF
par(mfrow = c(1, 2), mar = c(4, 4, 1, 1))
acf(deTrend, xlab = "Lag (in Years)", ylab = "Sample ACF", main = "")
pacf(deTrend, xlab = "Lag (in Years)", ylab = "Sample PACF", main = "")

# Histogram and QQ Plot
hist(deTrend, main = "", xlab = "", las = 1)
qqnorm(deTrend, main = "", pch = 16, cex = 0.8); qqline(deTrend, col = "blue")
```

### Model Selection and Fitting

```{r}
## AR(1)
(ar1.model <- arima(deTrend, order = c(1, 0, 0)))
ar1.resids <- resid(ar1.model)
plot(1875:1972, ar1.resids, type = "l", xlab = "Year", ylab = "AR1 Residuals")

## Sample ACF and PACF of the Residuals
par(mfrow = c(1, 2))
acf(ar1.resids, ylab = "Sample ACF", xlab = "Lag (in Years)")
pacf(ar1.resids, ylab = "Sample PACF", xlab = "Lag (in Years)")

## Normal Q-Q Plot for the Residuals
qqnorm(ar1.resids, main = ""); qqline(ar1.resids, col = "blue")

## Test for Time Dependence for the Residuals 
Box.test(ar1.resids, type = "Ljung-Box")

## AR(2)
(ar2.model <- arima(deTrend, order = c(2, 0, 0)))

## Calculate the Residuals
ar2.resids <- resid(ar2.model)

## Sample ACF and PACF of the Residuals
par(mfrow = c(1, 2))
acf(ar2.resids,  ylab = "Sample ACF", xlab = "Lag (in Years)")
pacf(ar2.resids,  ylab = "Sample PACF", xlab = "Lag (in Years)")

## Test for Time Dependence for the Residuals 
Box.test(ar2.resids, type = "Ljung-Box")

## Normal Q-Q Plot for the Residuals
qqnorm(ar2.resids, main = ""); qqline(ar2.resids, col = "blue")

## Fit the ARMA(2, 1) Model
(arma21.model <- arima(deTrend, order = c(2, 0, 1)))

## Calculate the Residuals
arma21.resids <- resid(arma21.model)

## Sample ACF and PACF of the Residuals
par(mfrow=c(1,2))
acf(arma21.resids,  ylab = "Sample ACF", xlab = "Lag (in Years)")
pacf(arma21.resids,  ylab = "Sample PACF", xlab = "Lag (in Years)")

## Normal Q-Q Plot for the Residuals
qqnorm(arma21.resids, main = ""); qqline(arma21.resids, col = "blue")

## Test
Box.test(arma21.resids, type = "Ljung-Box")

# Model Selection Using AIC
AIC(ar1.model); AIC(ar2.model); AIC(arma21.model)
```

### AR(2) Fitting and Forecasting

```{r, warning = FALSE}
# install.packages("forecast")
library(forecast)
(fit <- Arima(LakeHuron, order = c(2, 0, 0), include.drift = T))

par(mfrow = c(2, 2), mar = c(4.1, 4, 1, 0.8), las = 1)
res <- fit$residuals
plot(res, type = "l", xlab = "Year", ylab = "AR(2) Residuals", las = 1)
abline(h = 0, col = "blue")
qqnorm(res, main = ""); qqline(res, col = "blue")
acf(res,  ylab = "Sample ACF", xlab = "Lag (in Years)")
pacf(res,  ylab = "Sample PACF", xlab = "Lag (in Years)")

# 10-Year Forecasts
autoplot(forecast(fit, h = 10, level = c(50, 95)))
```