---
title: "Time Series Analysis 2"
author: "Blake Pappas"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: no
    toc_depth: 3
    fig_width: 8.5
    fig_height: 7
    fig_caption: yes
  word_document:
    toc: no
    toc_depth: '3'
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Seasonal Component Estimation

Now let's consider the situation that a time series consists of seasonal components only (assuming the trend has been estimated/removed). That is 
$$Y_{t} = s_{t} + \eta_{t}.$$
with $\{s_{t}\}$ having period $d$ (i.e., $s_{t} = s_{t+jd}$ for all integers $j$ and $t$). $\sum_{t = 1}^{d}s_{t} = 0$ and $\mathbb{E}[\eta_{t}] = 0.$ We can use a harmonic regression or a seasonal factor model to estimate the seasonal components or to use seasonal-differencing to remove the seasonality.

### Harmonic Regression

A harmonic regression model has the form

$$s_{t} = \sum_{j = 1}^{k}A_{k}\cos(2\pi f_{j} + \phi_{j}).$$ For each $j = 1, \cdots, k$:

*  $A_{j} > 0$ is the *amplitude* of the jth cosine wave. 

*  $f_{j}$ controls the the *frequency* of the j-th cosine wave (how often waves repeat). 

*  $\phi_{j} \in [-\pi, \pi]$ is the *phase* of the j-th wave (where it starts)

The above can be expressed as
$$\sum_{j = 1}^{k}\left(\beta_{1j}\cos(2\pi f_{j}) + \beta_{2j}\sin(2\pi f_{j})\right),$$
where $\beta_{1j} = A_{j}\cos(\phi_{j})$ and $\beta_{2j} = A_{j}\sin(\phi_{j}).$ Therefore, if the frequencies $\{f_{j}\}_{j = 1}^{k}$ are known, we can use regression techniques to estimate the parameters $\{\beta_{1j}, \beta_{2j}\}_{j = 1}^{k}$ by treating $\{\cos(2\pi f_{j})\}_{j = 1}^{k}$ and $\{\sin(2\pi f_{j})\}_{j = 1}^{k}$ as predictor variables. 

Let's use the monthly average temperature (in degrees Fahrenheit) recorded in Dubuque, IA from Jan. 1964 - Dec. 1975.

```{r, message = FALSE, warning = FALSE}
# install.packages("TSA")
library(TSA)
data(tempdub)
time <- as.numeric(time(tempdub))
par(mar = c(4, 4, 0.8, 0.6))
plot(time, tempdub, type = "l", las = 1, xlab = "", ylab = "")
points(time, tempdub, pch = 16, col = "blue", cex = 0.6)
grid()
mtext("Time (in Years)", side = 1, line = 2)
mtext(expression(paste("Monthly Average Temperature (", degree, "F)")), side = 2, line = 2)
```

First, we need to set up the harmonics (assuming yearly cycle):

```{r}
harmonics <- harmonic(tempdub, 1)
time <- as.numeric(time(tempdub))
par(mfrow = c(2, 1), las = 1, mar = c(2, 4, 0.8, 0.6))
plot(time, harmonics[, 1], type = "l", ylab = "cos")
plot(time, harmonics[, 2], type = "l", ylab = "sin")
```

Next, perform a linear regression using the harmonics as the predictors:

```{r}
harReg <- lm(tempdub ~ harmonics)
summary(harReg)

par(mar = c(3.6, 3.6, 0.8, 0.6))
plot(time, tempdub, type = "l", las = 1, xlab = "", ylab = "")
mtext("Time (in Years)", side = 1, line = 2)
mtext(expression(paste("Monthly Average Temperature (", degree, "F)")), side = 2, line = 2)
time <- as.numeric(time(tempdub))
lines(time, harReg$fitted.values, col = "blue", lty = 2)
abline(h = harReg$coefficients[1], lty = 2, col = "blue")
```

### Seasonal Factors

Harmonic regression assumes the seasonal pattern has a regular shape (i.e. the height of the peaks is the same as the depth of the troughs). Assuming the seasonal pattern repeats itself every $d$ time points, a less restrictive approach is to model it as 

$$s_{t} = \left\{ \begin{array}{ll}
         \beta_{1} & \mbox{for $t = 1, 1 + d, 1 + 2d, \cdots$};\\
        \beta_{2} & \mbox{for $t = 2, 2 + d, 2 + 2d,\cdots$};\\
        \vdots  & \vdots;\\
        \beta_{d} & \mbox{for $t = d, 2d, 3d,\cdots$}
        .\end{array} \right.$$
        
```{r}
month = season(tempdub)
season_means <- lm(tempdub ~ month - 1)
summary(season_means)

plot(time, tempdub, type = "l", las = 1, xlab = "", ylab = "")
mtext("Time (in Years)", side = 1, line = 2)
mtext(expression(paste("Monthly Average Temperature (", degree, "F)")), side = 2, line = 2)
points(time, season_means$fitted.values, col = "blue", pch = 16, cex = 0.8)
```

## Let's Put Trend and Seasonal Variation Together

Here we use the $\mathrm{CO}_{2}$ concentration time series as an example.
First, we perform a linear regression with both time and the harmonics as the covariates.

```{r}
time <- as.numeric(time(co2))
harmonics <- harmonic(co2, 1)

lm_trendSeason <- lm(co2 ~ time + harmonics)
summary(lm_trendSeason)

par(mar = c(3.8, 4, 0.8, 0.6))
plot(time, co2, type = "l", las = 1, xlab = "", ylab = "")
# points(co2, col = "blue", pch = 16, cex = 0.25)
mtext("Time (in Years)", side = 1, line = 2)
mtext(expression(paste("CO"[2], " Concentration (ppm)")), side = 2, line = 2.5)
grid()
lines(time, lm_trendSeason$fitted.values, col = "red", lty = 2)
```


## Read the Data

```{r}
airpass <- read.csv("airpass.csv", header = FALSE)

ts.plot(airpass)

## Create the Time Variable
yr <- seq(1949, 1960 + 11 / 12, by = 1 / 12)
```

## Plot the Time Series

```{r}
par(bty = "L", mar = c(3.6, 3.5, 0.8, 0.6), mgp = c(2.4, 1, 0), las = 1, mfrow = c(2, 1))

## Plot the Time Series
plot(yr, airpass$V1, type = "l", xlab = "Year",
     ylab = "Monthly Total (1000s)")
grid()

## Take a Log (To the Base 10) of the Air Passenger Data
log.airpass <- log10(airpass)
plot(yr, log.airpass$V1, type = "l", xlab = "Year", ylab = "log10(Monthly Total)")
grid()
```

## Plot Sample ACF and PACF

```{r}
log.shortair <- log.airpass$V1[1:132]
shortyears <- yr[1:132]
par(bty = "L", mar = c(3.6, 3.5, 0.8, 0.6), mgp = c(2.4, 1, 0), las = 1, mfrow = c(1, 2))
stats::acf(log.shortair, ylab = "Sample ACF", main = "", lag.max = 48, ylim = c(-0.6, 1), xaxt = "n")
axis(side = 1, at = seq(0, 48, 12))
stats::pacf(log.shortair, ylab = "Sample PACF", main = "", lag.max = 48, ylim = c(-0.6, 1), xaxt = "n")
axis(side = 1, at = seq(0, 48, 12))
```


## Trying Different Orders of Differencing

```{r}
## Take the Differences Y_t = (1-B) X_t
diff.1.0 <- diff(log.shortair)

## Take the Seasonal Differences Y_t =(1-B^(12)) X_t
diff.0.1 <- diff(log.shortair, lag = 12, diff = 1)

## Take the Differences Y_t = (1-B^(12)) (1-B) X_t
diff.1.1 <- diff(diff(log.shortair, lag = 12, diff = 1))

par(bty = "L", mar = c(3.6, 3.5, 1, 0.6), mgp = c(2.4, 1, 0), las = 1)
layout.matrix <- matrix(c(1, 1, 2, 3, 4, 4, 5, 6, 7, 7, 8, 9), nrow = 3, ncol = 4, byrow = T)
layout(mat = layout.matrix)
plot(shortyears[-1], diff.1.0, xlab = "", ylab = "d = 1, D = 0",
     type = "l", ylim = c(-0.1, 0.1), xlim = range(shortyears))

stats::acf(diff.1.0, lag.max = 48, ylab = "", xlab = "", main = "", ylim = c(-0.6, 1), xaxt = "n")
axis(side = 1, at = seq(0, 48, 12))
mtext("Sample ACF", side = 3, line = 0, cex = 0.8)

stats::pacf(diff.1.0, lag.max = 48, ylab = "", xlab = "", main = "", ylim = c(-0.6, 1), xaxt = "n")
axis(side = 1, at = seq(0, 48, 12))
mtext("Sample PACF", side = 3, line = 0, cex = 0.8)

plot(shortyears[-c(1:12)], diff.0.1, xlab = "", ylab = "d = 0, D = 1",
     type = "l", ylim = c(-0.1, 0.1), xlim = range(shortyears))

stats::acf(diff.0.1, lag.max = 48, ylab = "", xlab = "", main = "", ylim = c(-0.6, 1), xaxt = "n")
axis(side = 1, at = seq(0, 48, 12))
mtext("lag", side = 1, line = 1.8, cex = 0.8)

stats::pacf(diff.0.1, lag.max = 48, ylab = "", xlab = "", main = "", ylim = c(-0.6, 1), xaxt = "n")
axis(side = 1, at = seq(0, 48, 12))

plot(shortyears[-c(1:13)], diff.1.1, xlab = "", ylab = "d = 1, D = 1",
     type = "l", ylim = c(-0.1, 0.1), xlim = range(shortyears))
mtext("Year", side = 1, line = 1.8, cex = 0.8)

stats::acf(diff.1.1, lag.max = 48, ylab = "", xlab = "", main = "", ylim = c(-0.6, 1), xaxt = "n")
axis(side = 1, at = seq(0, 48, 12))
mtext("lag", side = 1, line = 1.8, cex = 0.8)

stats::pacf(diff.1.1, lag.max = 48, ylab = "", xlab = "", main = "", ylim = c(-0.6, 1), xaxt = "n")
axis(side = 1, at = seq(0, 48, 12))
mtext("lag", side = 1, line = 1.8, cex = 0.8)
```


## Show the ACF and PACF for the d = 1, D = 0 Case

```{r}
par(mfrow = c(1, 2), cex = 0.8, bty = "L", mar = c(3.6, 3, 1, 0.6), mgp = c(2.4, 1, 0), las = 1)
stats::acf(diff.1.0, lag.max = 48, ylab = "", xlab = "", main = "", ylim = c(-0.6, 1), xaxt = "n")
axis(side = 1, at = seq(0, 48, 12))
mtext("Sample ACF", side = 3, cex = 0.8)

stats::pacf(diff.1.0, lag.max = 48, ylab = "", xlab = "", main = "", ylim = c(-0.6, 1), xaxt = "n")
axis(side = 1, at = seq(0, 48, 12))
mtext("Sample PACF", side = 3, cex = 0.8)
```

## A Useful Function for the Model Diagnostics (Courtesy of Peter Craigmile at OSU)

```{r}
plot.residuals <- function (x, y = NULL, lag.max = NULL, mean.line = TRUE,
			    acf.ylim = c(-0.25, 1), mfrow = c(2, 2),
                            lags = NULL, ...) {
  if (!is.null(mfrow))
    par(mfrow = mfrow)
  if (is.null(y)){
    y <- x
    x <- seq(length(y))
  } else {
    x <- as.numeric(x)
    y <- as.numeric(y)
  }

  if (is.null(lag.max)) {
    lag.max <-  floor(10 * log10(length(x)))
  }
  plot(x, y, type = "l", ...)
  if (mean.line) abline(h = 0, lty = 2)
  qqnorm(y, main = "", las = 1); qqline(y)
  if (is.null(lags)) {
    acf(y, main = "", lag.max = lag.max, xlim = c(0, lag.max), ylim = acf.ylim,
        ylab = "Sample ACF", las = 1)

    pacf(y, main = "", lag.max = lag.max, xlim = c(0, lag.max), ylim = acf.ylim,
         ylab = "Sample PACF", las = 1)
  }  
  else {
    stats::acf(y, main = "", lag.max = lag.max, xlim = c(0, lag.max), ylim = acf.ylim,
        ylab = "Sample ACF", xaxt = "n", las = 1)
    axis(side = 1, at = lags)

    stats::pacf(y, main = "", lag.max = lag.max, xlim = c(0, lag.max), ylim = acf.ylim,
         ylab = "Sample PACF", xaxt = "n", las = 1)
    axis(side = 1, at = lags)
  }
  Box.test(y, lag.max, type = "Ljung-Box")
}
```


## Fitting the SARIMA$(1, 1, 0)\times(1, 0, 0)$ Model

```{r}
fit1 <- arima(diff.1.0, order = c(1, 0, 0), seasonal = list(order = c(1, 0, 0), period = 12))
fit1
Box.test(fit1$residuals, lag = 48, type = "Ljung-Box")
par(mfrow = c(2, 2), cex = 0.8, bty = "L", mar = c(3.6, 4, 0.8, 0.6),
    mgp = c(2.8, 1, 0), las = 1)
plot.residuals(shortyears[-1], resid(fit1), lag.max = 48,
               ylab = "SARIMA Residuals", xlab = "Year", lags = seq(0, 48, 12))
```

## Fitting the SARIMA$(0, 1, 0)\times(1, 0, 0)$ Model


```{r}
(fit2 <- arima(diff.1.0, seasonal = list(order = c(1, 0, 0), period = 12)))
Box.test(fit2$residuals, lag = 48, type = "Ljung-Box")
par(mfrow = c(2, 2), cex = 0.8, bty = "L", mar = c(3.6, 4, 0.8, 0.6),
    mgp = c(2.8, 1, 0), las = 1)
plot.residuals(shortyears[-1], resid(fit2), lag.max = 48,
               ylab = "SARIMA Residuals", xlab = "Year", lags = seq(0, 48, 12))
```

## Forecasting the 1960 Data

```{r}
## Fit the First Full Model
fit1 <- arima(log.shortair, order = c(1, 1, 0),
                     seasonal = list(order = c(1, 0, 0), period = 12))
fit1

## Fit the Second Full Model
fit2 <- arima(log.shortair, order = c(0, 1, 0),
                     seasonal = list(order = c(1, 0, 0), period = 12))
fit2

## Define the Forecasting Time Points
fyears <- yr[133:144]

preds1 <- predict(fit1, 12)
forecast1 <- preds1$pred
flimits1  <- qnorm(0.975) * preds1$se

preds2 <- predict(fit2, 12)
forecast2 <- preds2$pred
flimits2  <- qnorm(0.975) * preds2$se

par(mfrow = c(2, 2), cex = 0.8, bty = "L", mar = c(3.6, 4, 1, 0.6),
    mgp = c(2.4, 1, 0), las = 1)

plot(shortyears, log.shortair, type = "l", xlab = "Year",
     ylab = "log10(Passenger Numbers)", xlim = range(yr), ylim = c(2, 2.9))
mtext("SARIMA(1,1,0) x (1,0,0)")

## Plot the Forecasts
lines(fyears, forecast1, lwd = 2, col = "blue")

## Plot the 95% Prediction Intervals
lines(fyears, forecast1 + flimits1, lty = 3, col = "blue")
lines(fyears, forecast1 - flimits1, lty = 3, col = "blue")

plot(shortyears, log.shortair, type = "l", xlab = "Year",
     ylab = "log10(Passenger Numbers)", xlim = range(yr), ylim = c(2, 2.9))
mtext("SARIMA(0,1,0) x (1,0,0)")

## Plot the Forecasts
lines(fyears, forecast2, lwd = 2, col = "red")

## Plot the 95% Prediction Intervals
lines(fyears, forecast2 + flimits2, lty = 3, col = "red")
lines(fyears, forecast2 - flimits2, lty = 3, col = "red")

plot(shortyears, 10^log.shortair, type = "l", xlab = "Year",
     ylab="1000s of Airline Passengers", xlim = range(yr), ylim = c(100, 800))
lines(fyears, 10^forecast1, lwd = 2, col = "blue")
lines(fyears, 10^(forecast1 + flimits1), lty = 3, col = "blue")
lines(fyears, 10^(forecast1 - flimits1), lty = 3, col = "blue")

plot(shortyears, 10^log.shortair, type = "l", xlab = "Year",
     ylab="1000s of Airline Passengers", xlim = range(yr), ylim = c(100, 800))
lines(fyears, 10^forecast2, lwd = 2, col = "red")
lines(fyears, 10^(forecast2 + flimits2), lty = 3, col = "red")
lines(fyears, 10^(forecast2 - flimits2), lty = 3, col = "red")
```

## Evaluating Forecast Performance

```{r}
## Calculate the Root Mean Square Error (RMSE)
sqrt(mean((10^forecast1 - 10^log.airpass$V1[133:144])^2))
sqrt(mean((10^forecast2 - 10^log.airpass$V1[133:144])^2))

## Calculate the Mean Relative Prediction Error
mean((10^forecast1 - 10^log.airpass$V1[133:144]) / 10^log.airpass$V1[133:144])
mean((10^forecast2 - 10^log.airpass$V1[133:144]) / 10^log.airpass$V1[133:144])

## Calculate the Empirical Coverage Rate
PI_fit1 <- cbind(as.numeric(10^(forecast1 + flimits1)),
                 as.numeric(10^(forecast1 - flimits1)))
sum(apply(PI_fit1 - 10^log.airpass$V1[133:144], 1, prod) < 0) / length(10^log.airpass$V1[133:144])
PI_fit2 <- cbind(as.numeric(10^(forecast2 + flimits2)),
                 as.numeric(10^(forecast2 - flimits2)))
sum(apply(PI_fit2 - 10^log.airpass$V1[133:144], 1, prod) < 0) / length(10^log.airpass$V1[133:144])
```