---
title: 'Completely Randomized Designs - Lab'
author: "Blake Pappas"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The dataset \texttt{PlantGrowth} contains results from an experiment to compare yields (as measured by dried weight of plants) obtained under a control and two different treatment conditions.

*Data Source:* Dobson, A. J. (1983) *An Introduction to Statistical Modelling*. London: Chapman and Hall.

Let's load the data first:

**Code:**

```{r}
data(PlantGrowth)
str(PlantGrowth)
```

1. Compute the treatment means and standard deviations.

**Code:**

```{r}
(means <- tapply(PlantGrowth$weight, PlantGrowth$group, mean))
(vars <- tapply(PlantGrowth$weight, PlantGrowth$group, sd))
```

2. Make side-by-side boxplots by treatment.

**Code:**

```{r}
boxplot(weight ~ group, data = PlantGrowth, las = 1)
boxplot(weight ~ group, data = PlantGrowth, las = 1, horizontal = T)
```

3. Write down the effects model and explain each term in the model (including the model assumptions regarding the random error).

**Answer: $Y_{ij}: \mu + \alpha_{i} + \epsilon_{ij}$, i = 1,...,g,   j = 1,..., $n_{i}$,   $\epsilon_{ij}$ ~ N(0,$\sigma$ ^2). $Y_{ij}$ is the random variable that represents the response for the jth experimental unit to treatment i. $\mu_{i}$ is the mean response time for the ith treatment. $\epsilon_{ij}$ is the random error, which is assumed to be normal.**


4. Perform an overall F-test using ANOVA. State the hypotheses, p-value, decision, and conclusion.

**Code:**

```{r}
AOV <- aov(weight ~ group, data = PlantGrowth)
summary(AOV)
```

**Answer: The null hypothesis is $H_{0}: \alpha_{\texttt{i}}= 0$, the alternative hypothesis is $H_{a}: \alpha_{\texttt{i}}\neq 0$, the $\alpha$ is 0.05, the F-statistic is 4.846, the p-value is 0.0159, and the decision is to reject the null hypothesis. The data shows evidence of differences the three treatment groups.**


5. Conduct pairwise comparisons using Tukey's HSD  procedure.

**Code:**

```{r}
# HSD
HSD <- TukeyHSD(AOV, conf.level = 0.95)
HSD$group
```

**Answer: The trt2-trt1 groups have statistically significant differences.**

6. Use qqplot to examine the normality assumption on error.

**Code:**

```{r}
mod1 <- lm(weight ~ group, data = PlantGrowth)

qqnorm(mod1$resid, cex = 0.8)
qqline(mod1$resid, col = "red", lwd = 1.5)
```

**Answer: Based on the plot, the modeling assumption of normality appears to be reasonable. The pattern of the plot runs very close to the trend line.**

7. Make a residual plot to assess equal variance assumption.

**Code:**

```{r}
plot(mod1$fitted, mod1$resid, las = 1, xlab = "Fitted Value", ylab = "Residual", cex = 0.75, col = "blue")
abline(h = 0)

plot(mod1$resid, col = "blue", las = 1, cex = 0.75, pch = 16, ylab = "Residual")
abline(h = 0, lty = 2, lwd = 1.5)

# By Treatment
attach(PlantGrowth)
colorf <- as.factor(PlantGrowth$group)
colors <- c("green",
"orange", "blue")

plot(mod1$resid, col = colors[colorf], las = 1, cex = 0.75,
     pch = 16, ylab = "Residual")
abline(h = 0, lty = 2, lwd = 1.5)
```

**Answer: Based on the residual plots, the assumption of equal variances seems to be reasonable.**

8. Perform a Levene’s test for equal variance.

**Code:**

```{r, warning = FALSE}
# Levene’s Test for Equal Variance
# install.packages("lawstat")
library("lawstat")
levene.test(PlantGrowth$weight, PlantGrowth$group, location = "mean")

# Brown-Forsythe Test
levene.test(PlantGrowth$weight, PlantGrowth$group, location = "median")
```