---
title: "R Basics"
author: "Blake Pappas"
date: "`r Sys.Date()`"
output: pdf_document
bookmarks: false
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Data cleaning and visualization in R
# We will use the "coffee.csv" dataset to practice.

# Load the dplyr package:
```{r, warning = FALSE, message = FALSE}
# install.packages("dplyr")

library(dplyr)
```

# Import the "coffee.csv" dataset:
```{r}
coffee = read.csv("coffee.csv", sep = ";")
```

# P1: Arrange the data
# What is the ProductID of the most profitable coffee?
# Answer: The ProductID of the most profitable coffee is 2.
# Write the code below:
```{r}
most_profitable = coffee %>% arrange(desc(Profit))

head(most_profitable)
```

# P2: Subset the data
# Find the subset of data with "Sales" larger than 200.
# Write the code below:
```{r}
sales200 = coffee %>% filter(Sales > 200)

head(sales200)
```

# P3: Group and summarize the data
# Find out the total profit and average inventory level of each product (identified by a unique "ProductID").
```{r}
data_group = coffee %>% group_by(ProductId) %>%
  summarise(total_profit = sum(Profit),
            average_inventory_level = mean(Inventory))

data_group
```

# P4: Create a new variable, "ProfitRatio", which equals "Profit"/"Sales"
```{r}
data_ProfitRatio = coffee %>% mutate(ProfitRatio = Profit / Sales)

head(data_ProfitRatio)
```

# P5: Make a histogram
# Plot the distribution of "COGS".
# Write the code below:
``` {r}
hist(coffee$COGS)
```

# P6: Make a boxplot
# Plot the distribution of "Profit" across different "ProductIDs".
# Write the code below:
```{r}
boxplot(Profit ~ ProductId, data = coffee)
```

# P7: Make a scatterplot
# Plot the relationship between "Margin" and "Inventory".
# Write the code below:
```{r}
plot(coffee$Margin, coffee$Inventory)
```