---
title: "Continuous Probability Distributions and Sampling Distributions"
author: "Blake Pappas"
date: "10/5/2021"
output:
  pdf_document: default
  word_document: default
  html_document: default
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Normal Quantile Plots

Very often in statistics we assume that observed data values are realizations of random variables from a normal distribution. A normal quantile plot is used to check whether that assumption is reasonable.  

If data are from a perfect normal distribution, the normal quantile plot will show points along a perfect 45-degree, uphill line. The code below will generate random normal variables and display the quantile plot.

```{r, warning = FALSE, message = FALSE}
x1 <- rnorm(5e3)
qqnorm(x1)
qqline(x1, col = 'red')  # Adds the line that perfectly normal data would follow
```

With a smaller sample, the plot will not look as perfect, but the general linear pattern will be the same. 

```{r}
x2 <- rnorm(15)
qqnorm(x2)
qqline(x2, col = 'red')  
```

If data are not normal, you may see an S-shaped pattern in the quantile plot, or curvature in the lower or upper tail. The chunks below generate data from some skewed distributions and show their quantile plots.


```{r}
par(mfrow = c(1, 2))
x3 <- rgamma(5e3, 1, 10)  # Samples from a very skewed gamma distribution
hist(x3, main = 'x3 - not normal') 
qqnorm(x3)
qqline(x3, col = 'red')
```

```{r}
par(mfrow = c(1, 2))
x4 <- rgamma(5e3, 10, 10) # Samples from a nearly symmetric gamma distribution
hist(x4, main = 'x4 - not normal') 
qqnorm(x4)
qqline(x4,col = 'red')
```

```{r}
par(mfrow = c(1, 2))
x5 <- rpois(5e3, 4) # Samples from a discrete distribution
hist(x5, main = 'x5 - not normal \n discrete') 
qqnorm(x5)
qqline(x5,col = 'red')
```

In some of these examples, the histograms clearly indicate that the shape of the distribution does not resemble a normal curve. Sometimes, however, the quantile plot reveals patterns that are harder to spot in a histogram. Here is a sample from a t-distribution, which is bell-shaped but has fatter tails than a normal curve. The non-normality is much easier to spot in the normal quantile plot.

```{r}
par(mfrow = c(1, 2))
x6 <- rt(5e3, 6)
hist(x6, breaks = 20)
qqnorm(x6)
qqline(x6,col = 'red')
```

## Exercises

## Exercise 1: Normal Probabilities
Use the functions `pnorm()` and `qnorm()` to find the following probabilities.

a. Find the 77th percentile of a normal distribution with mean 0 and standard deviation of 0.5.

```{r}
qnorm(0.77, 0, 0.5)
```

**Answer: The 77th percentile of the distribution is approximately 0.3694.**

b. Find the probability that a N(0, 0.5^2) random variable is less than -0.35.

```{r}
pnorm(-0.35, 0, 0.5)
```

**Answer: The probability that N is less than -0.35 is approximately 0.2420.**

c. Find the probability that a N(10, 3^2) random variable is greater than 17.

```{r}
pnorm(17, 10, 3, lower.tail = FALSE)
```

**Answer: The probability that N is greater than 17 is approximately 0.0100.**

d. Find the probability that a N(10, 3^2) random variable is between 9 and 14.

```{r}
pnorm(14, 10, 3) - pnorm(9, 10, 3)
```

**Answer: The probability that N is between 9 and 14 is approximately 0.5393.**


## Exercise 2: Normal Probabilities
The age of employees at a certain company have a Normal(37, 4.7^2) distribution.

a. Find the probability that a randomly selected employee is older than 40.

```{r}
pnorm(40, 37, 4.7, lower.tail = FALSE)
```

**Answer: The probability that a randomly selected employee is older than 40 is approximately 0.2616.**

b. Find the cutoff for the youngest 10% of employees; that is, 10% of employees are younger than what age?

```{r}
qnorm(0.1, 37, 4.7)
```

**Answer: 10% of employees are younger than the age of 30.9767.**

c. Find the probability that a randomly selected employee is between 33 and 39.

```{r}
pnorm(39, 37, 4.7) - pnorm(33, 37, 4.7)
```

**Answer: The probability that a randomly selected employee is between 33 and 39 is approximately 0.4674.**

d. Find the cutoff for the oldest 5% of employees; that is, 5% of employees are older than what age?

```{r}
qnorm(0.95, 37, 4.7)
```

**Answer: 5% of employees are older than 44.7308 years.**

## Exercise 3: Simulation to Approximate Probabilities
Simulation can be used to find approximate probabilities. It can also be used to find approximate distributions. If you take a large number of samples (no less than 500) from a probability distribution, the histogram of the samples will look very similar to the true probability density function (or pmf). 

a. The `rnorm()` function can be used to take samples from a normal distribution. Here is an example of using `rnorm` to draw n = 5 samples from a N(-4, 0.8^2) distribution.

```{r}
y <- rnorm(n = 5, mean = -4, sd = 0.8)
print(y)
```

Now use `rnorm` to draw 1000 samples from the N(21.4, 2.8^2) distribution. Make a histogram of the samples. Does the shape of the histogram resemble a normal curve? Then calculate the mean and standard deviation of the samples. Do these numbers come close to the true mean and standard deviation?

```{r}
y <- rnorm(n = 1000, mean = 21.4, sd = 2.8)
hist(y)
```

```{r}
mean(y)
```

```{r}
sd(y)
```

**Answer: Yes, the shape of the histogram does resemble that of a normal curve. Yes, these numbers do come close to the true mean and standard deviation. However, they are not exact. A larger sample size would bring them closer to the true mean and standard deviation.**

b. Using the samples from part (a), calculate the proportion of samples that are greater than 25. Compare this to the true probability that a N(21.4, 2.8^2) random variable is greater than 25, which you can find using `pnorm()`. 

```{r}
pnorm(25, mean(y), sd(y), lower.tail = FALSE)
```

```{r}
pnorm(25, 21.4, 2.8, lower.tail = FALSE)
```

**Answer: The sample's probability is less than the true probability. Again, a larger sample size would bring this calculated probability closer to the true probability.**

c. The code below gives an example of drawing random samples from a gamma distribution using the `rgamma` function. The code draws n = 5 samples from the Gamma distribution with shape parameter equal to 0.5 and scale parameter equal to 1. (This is sometimes called the Gamma(0.5, 1) distribution.) 

```{r}
y <- rgamma(5, shape = 0.5, scale = 1)
print(y)
```

Modify the code to generate n = 10000 samples. Create a histogram and comment on the shape of the distribution. This shape will be similar to the true density curve of the Gamma(0.5, 1) distribution.

```{r}
y <- rgamma(10000, shape = 0.5, scale = 1)
hist(y)
```

```{r}
mean(y)
```

```{r}
median(y)
```

**Answer: The shape of the gamma distribution is unimodal and extremely right-skewed.**

d. Using the samples you obtained in part c, approximate the probability that a Gamma(0.5, 1) random variable is between 1 and 2.

```{r}
pnorm(2, mean(y), sd(y)) - pnorm(1, mean(y), sd(y))
```

**Answer: The probability that a random variable is between 1 and 2 is approximately 0.22.**

## Exercise 4: Quantile Plots

a. Use the `rnorm` function to generate a sample of size 100 from a Normal(4, 0.75^2) distribution. Create a normal quantile plot of the sample and state what features of the plot indicate whether it is reasonable to consider the sample to be approximately normal.
```{r}
x1 <- rnorm(n = 100, mean = 4, sd = 0.75)
qqnorm(x1)
qqline(x1, col = 'red')
```

**Answer: It is not reasonable to consider the sample to be approximately normal because the plotted points do not seem to follow along a 45-degree, uphill line. Instead, they follow an S-shaped pattern along the trend line.**

b. Use the `rgamma()` function to generate a sample of size 75 from a Gamma distribution with shape = 0.5 and scale = 1. Create a normal quantile plot of the sample and state what features of the plot indicate whether it is reasonable to consider the sample to be approximately normal.

```{r}
x1 <- rgamma(75, shape = 0.5, scale = 1)
qqnorm(x1)
qqline(x1, col = 'red')
```

**Answer: It is also not reasonable to consider this sample to be approximately normal because of the skewed distributions in the plotting.**