---
title: "Multivariate Linear Regression"
author: "Blake Pappas"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: false
    toc_depth: 3
    fig_width: 6
    fig_height: 5
    fig_caption: yes
header-includes:
   - \usepackage{animate}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the Motor Trend Car Road Tests Data

```{r}
data(mtcars)
head(mtcars)
```

## Data Manipulation

```{r}
mtcars$cyl <- factor(mtcars$cyl)
vars <- c("mpg", "disp", "hp", "wt") # Response variables
Y <- as.matrix(mtcars[, vars])
```

## Summarizing the Responses

```{r, warning = FALSE}
colMeans(Y) # Calculates the means of the four response variables
apply(Y, 2, sd) # Calculates the standard deviation of the four response variables
cov(Y) # Calculates the covariance of the four response variables
cor(Y) # Calculates the correlation matrix for the four response variables
library(GGally)
ggpairs(as.data.frame(Y))
```

## Fitting Linear Regression

```{r}
mvlm <- lm(Y ~ cyl + am + carb, data = mtcars) # Creates a multiple linear regression model between the four response variables (Y) and the three predictor variables
summary(mvlm)
```

## Sum Squared Cross Product (SSCP)

```{r}
ybar <- colMeans(Y)
n <- nrow(Y); d <- ncol(Y)
Ybar <- matrix(ybar, n, d, byrow = TRUE)
SSCP.Tot <- crossprod(Y - Ybar)
SSCP.Reg <- crossprod(mvlm$fitted.values - Ybar)
SSCP.Err <- crossprod(Y - mvlm$fitted.values)
SSCP.Tot
SSCP.Reg + SSCP.Err
```

## Estimated Error Covariance Matrix in R

```{r}
p <- nrow(coef(mvlm)) - 1
SigmaHat <- SSCP.Err / (n - p - 1)
SigmaHat
```

## Testing If We Need "cyl"

```{r}
# Tests Whether the Reduced Model Works Better Than the Full Model
mvlm0 <- lm(Y ~ am + carb, data = mtcars)
anova(mvlm, mvlm0, test = "Wilks")
anova(mvlm, mvlm0, test = "Pillai")
```

## Confidence and Prediction Intervals

```{r}
newdata <- data.frame(cyl = factor(6, levels = c(4, 6, 8)),
                      am = 1, carb = 4)

# Confidence Interval
predict(mvlm, newdata, interval = "confidence")

# Prediction Interval
predict(mvlm, newdata, interval = "prediction")
```

R does not yet have the capability to produce CIs/PIs with multivariate responses.

Below is the R function used to calculate the multivariate regression CIs and PIs:

```{r}
pred.mlm <- function(object, newdata, level = 0.95,
                     interval = c("confidence", "prediction")) {
  form <- as.formula(paste("~", as.character(formula(object))[3]))
  xnew <- model.matrix(form, newdata)
  fit <- predict(object, newdata)
  Y <- model.frame(object)[, 1]; X <- model.matrix(object)
  n <- nrow(Y); d <- ncol(Y); p <- ncol(X) - 1
  sigmas <- colSums((Y - object$fitted.values)^2) / (n - p - 1)
  fit.var <- diag(xnew %*% tcrossprod(solve(crossprod(X)), xnew))
  if(interval[1] == "prediction") fit.var <- fit.var + 1
  const <- qf(level, df1 = d, df2 = n - p - d) * d * (n - p - 1) / (n - p - d)
  vmat <- (n / (n - p - 1)) * outer(fit.var, sigmas)
  lwr <- fit - sqrt(const) * sqrt(vmat)
  upr <- fit + sqrt(const) * sqrt(vmat)
  if(nrow(xnew) == 1L) {
    ci <- rbind(fit, lwr, upr)
    rownames(ci) <- c("fit", "lwr", "upr")
  } else {
    ci <- array(0, dim = c(nrow(xnew), d, 3))
    dimnames(ci) <- list(1:nrow(xnew), colnames(Y), c("fit", "lwr", "upr"))
    ci[, , 1] <- fit; ci[, , 2] <- lwr; ci[, , 3] <- upr
  }
  ci
}

# Confidence Interval
pred.mlm(mvlm, newdata)

# Prediction Interval
pred.mlm(mvlm, newdata, interval = "prediction")
```