---
title: 'Simple Linear Regression'
author: "Blake Pappas"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  pdf_document:
    toc: false
    toc_depth: 3
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Example: Maximum Heart Rate vs. Age

The maximum heart rate ($\text{HR}_{max}$) of a person is often said to be related to age (Age) by the equation:
$$\text{HR}_{max} = 220 - \text{Age}$$

Let's use a dataset to assess this statement.


## Load the Dataset

```{r}
dat <- read.csv('maxHeartRate.csv', header = T) 
head(dat)
```


## Examine the Data Before Fitting Models

```{r}
y <- dat$MaxHeartRate; x <- dat$ï..Age
summary(dat)
var(x); var(y)
cov(x, y)
cor(x, y)
```

## Plot the Data Before Fitting Models

This is what the scatterplot would look like by default. Put predictor (age) to the first argument and response (maxHeartRate) to the second argument.


```{r}
plot(x, y)
```

Let's make the plot look nicer (type ?plot to learn more).

```{r}
par(las = 1, mar = c(4.1, 4.1, 1.1, 1.1))
plot(x, y, pch = 16, xlab = "Age (Years)", ylab = "Maximum Heart Rate (bpm)")
grid()
```


# Simple Linear Regression

## Estimation

Let's do the calculations to figure out the regression coefficients, as well as the standard deviation of the random error.   

### Slope: $\hat{\beta}_{1} = \frac{\sum_{i = 1}^{n}(y_{i} - \bar{y})(x_{i} - \bar{x})}{\sum_{i = 1}^{n}(x_{i} - \bar{x})^2}$

```{r}
y_diff <- y - mean(y)
x_diff <- x - mean(x)
beta_1 <- sum(y_diff * x_diff) / sum((x_diff)^2)
beta_1
```

### Intercept: $\hat{\beta}_{0}= \bar{y} - \bar{x} \hat{\beta}_{1}$

```{r}
beta_0 <- mean(y) - mean(x) * beta_1
beta_0
```

### Fitted Values: $\hat{y} = \hat{\beta}_{0} + \hat{\beta}_{1}x$

```{r}
y_hat <- beta_0 + beta_1 * x
y_hat
```

### $\hat{\sigma}$: $\hat{\sigma}^2 = \frac{\sum_{i = 1}^{n}(y_{i} - \hat{y}_{i})^2}{n - 2}$

```{r}
sigma2 <- sum((y - y_hat)^2) / (length(y) - 2)
sqrt(sigma2)
```

Add the fitted regression line to the scatterplot:

```{r}
par(las = 1, mar = c(4.1, 4.1, 1.1, 1.1))
plot(x, y, pch = 16, xlab = "Age", ylab = "Maximum Heart Rate (bpm)")
grid()
abline(a = beta_0, b = beta_1, col = "red")
```

Let R do all the work:

```{r}
fit <- lm(MaxHeartRate ~ ï..Age, data = dat)
summary(fit)
```

* Regression coefficients

```{r}
fit$coefficients
```

* Fitted values

```{r}
fit$fitted.values
```

* $\hat{\sigma}$

```{r}
summary(fit)$sigma
```

## Model Checking

### Residual Plots

```{r}
## Residuals vs. x
par(las = 1, mar = c(4.1, 4.1, 1.1, 1.1))
plot(x, fit$residuals, pch = 16, ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)

## Residuals vs. yhat
par(las = 1, mar = c(4.1, 4.1, 1.1, 1.1))
plot(fit$fitted.values, fit$residuals, pch = 16, ylab = "Residuals", xlab = expression(hat(y)))
abline(h = 0, col = "red", lty = 2)
```

### Assessing Normality of Random Error

```{r}
# Histogram
hist(fit$residuals, col = "lightblue", border = "gray", las = 1)

# qqplot
qqnorm(fit$residuals, pch = 16, las = 1)
qqline(fit$residuals)
```


## Statistical Inference

### Confidence Intervals for $\beta_{0}$ and $\beta_{1}$

```{r}
alpha = 0.05
beta1_hat <- summary(fit)[["coefficients"]][, 1][2]
se_beta1 <- summary(fit)[["coefficients"]][, 2][2]
CI_beta1 <- c(beta1_hat - qt(1 - alpha / 2, 13) * se_beta1,
              beta1_hat + qt(1 - alpha / 2, 13) * se_beta1)
CI_beta1
 
confint(fit)
# confint(fit, level = 0.9)  -- option to change confidence level
```

### Confidence and Prediction Intervals for $\mathrm{E}[Y_{new}|x_{new} = 40]$

```{r}
Age_new = data.frame(ï..Age = 40)
hat_Y <- fit$coefficients[1] + fit$coefficients[2] * 40
hat_Y
predict(fit, Age_new, interval = "confidence", level = 0.9) 
predict(fit, Age_new, interval = "predict", level = 0.95)
```


### Check

```{r}
sd <- sqrt((sum(fit$residuals^2) / 13))
ME <- qt(1 - alpha / 2, 13) * sd * sqrt(1 + 1 / 15 + (40 - mean(x))^(2) / sum((x - mean(x))^2))
c(hat_Y - ME, hat_Y + ME)
```

### Constrcuting Pointwise CIs and PIs

```{r, warning = FALSE}
Age_grid = data.frame(ï..Age = 18:72)
CI_band <- predict(fit, Age_grid, interval = "confidence") 
PI_band <- predict(fit, Age_grid, interval = "predict")

plot(dat$ï..Age, dat$MaxHeartRate, pch = 16, cex = 0.75,
     xlab = "Age (Years)", ylab = "Maximum Heart Rate (bpm)", las = 1)
abline(fit, col = "blue")
abline(v = mean(dat$Age), lty = 2, col = "gray")
abline(h = mean(dat$MaxHeartRate), lty = 2, col = "gray")
lines(18:72, CI_band[, 2], lty = 2, col = "blue")
lines(18:72, CI_band[, 3], lty = 2, col = "blue")
lines(18:72, PI_band[, 2], lty = 2, col = "darkgreen")
lines(18:72, PI_band[, 3], lty = 2, col = "darkgreen")
legend("topright", legend = c("CI", "PI"), col = c("blue", "darkgreen"), lty = 2, bty = "n")
```


### Hypothesis Tests for $\beta_{1}$

$H_{0}: \beta_{1} = -1$ vs. $H_{a}: \beta_{1} \neq -1$ with $\alpha = 0.05$

```{r}
beta1_null <- -1
t_star <- (beta1_hat - beta1_null) / se_beta1
p_value <- 2 * pt(t_star, 13, lower.tail = F) 
p_value

par(las = 1)
x_grid <- seq(-3.75, 3.75, 0.01)
y_grid <- dt(x_grid, 13)
plot(x_grid, y_grid, type = "l", xlab = "Test Statistic", ylab = "Density", xlim = c(-3.75, 3.75))
polygon(c(x_grid[x_grid < -t_star], rev(x_grid[x_grid < -t_star])),
        c(y_grid[x_grid < -t_star], rep(0, length(y_grid[x_grid < -t_star]))), col = "skyblue")

polygon(c(x_grid[x_grid > t_star], rev(x_grid[x_grid > t_star])),
        c(y_grid[x_grid > t_star], rep(0, length(y_grid[x_grid > t_star]))), col = "skyblue")
abline(v = t_star, lty = 2)
abline(v = -t_star, lty = 2)
abline(h = 0)
```